---
title: Inaction on climate change projected to reduce European life expectancy

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Mathew E. Hauer
  # thanks: The authors gratefully acknowledge ...
  affiliation: Department of Sociology, Florida State University

- name: Alexis R. Santos
  affiliation: Human Development and Family Studies, Pennsylvania State University

keywords:
- climate change
- life expectancy
- mortality
- Europe


abstract: |
  Climate change-related excess mortality estimates clearly demonstrate a dramatic impact on public health and human mortality. However, life expectancy at birth is more easily communicated and understood by the public. By properly situating climate change mortality within the contexts of life expectancy, we better represent the cost of climate change on longevity. In this paper, we convert excess mortality estimates due to increases in extreme weather from climate change into potential reductions in life expectancy at birth in thirty-one European countries. We project climate change extremes to reduce life expectancy at birth by 0.24 years for the average European country with differences in excess of 1.0 years in some countries by the end of the century. We only estimate the impact of mortality directly related to climate extremes, making our estimates conservative. Thus, the cost of inaction on climate change could approach, and likely to exceed, one year of life in some European countries.

bibliography: ccmort.bib
output: rticles::asa_article
header-includes:
  - \usepackage{rotating}
  - \usepackage[table]{xcolor}
  - \usepackage{tabularx}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---


```{r load, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
read_chunk('../MANUSCRIPT/R/SCRIPTS/0001-libraries.R')
# read_chunk('../MANUSCRIPT/R/SCRIPTS/figure-GBD-death-distribution.R')
read_chunk('../MANUSCRIPT/R/SCRIPTS/003-calculate-cc-e0.R')
read_chunk('../MANUSCRIPT/R/SCRIPTS/001-format-odds-and-ends.R')
read_chunk('../MANUSCRIPT/R/SCRIPTS/101-figure-lollipop.R')
read_chunk('../MANUSCRIPT/R/SCRIPTS/102-figure-map.R')
```

```{r libraries, include = FALSE}
```

```{r format-odds-and-ends, include = FALSE}
```

```{r calculate-cc-e0, include = FALSE}
```


# Main Text

Climate change's implications on humanity go far beyond estimates of economic damages [@hsiang2017estimating], estimates of displacement [@rigaud2018groundswell], or human conflict [@barnett2007climate]  but have the potential to contribute to the loss of human life [@forzieri2017increasing;@pachauri2014climate]. As impact quantification studies move further into the social sciences, properly quantifying and conveying the impact of climate change on public health is of increasing importance [@melillo2014climate;@cloyd2016engagement].

Scholars have long estimated the mortality risks associated with climate change and typically use excess or extra mortality [@forzieri2017increasing;@wilson2017climate;@mcmichael2006climate;@zanobetti2012summer]. For example, Forzieri et al. [-@forzieri2017increasing] project climate change extremes to increase European mortality by 150,000+ deaths per year by the end of the century. Although such estimates are useful, the excess mortality estimates are rather sterile -- one death is a tragedy, a million deaths a statistic -- and difficult to relate to on a personal level. Life expectancy at birth ($e_0$), on the other hand, provides potent comparisons of mortality vectors and converts excess mortality into an intuitively understood metric, relatable to everyone. To our knowledge, no studies exist that comprehensively examine the potential reductions of life expectancy due to climate change. Without properly situating the potential loss of life within the contexts of metrics such as life expectancy, we risk underestimating and misreporting the impact of climate change on human mortality.

Using published data on excess mortality [@forzieri2017increasing], we connect climate change excess mortality to life expectancy in a mortality model. By estimating the increase in age-specific mortality rates associated with previously published excess death estimates, we can assess how much climate change could reduce the anticipated longevity of the average person in thirty-one European countries. This approach allows us to quantify the impact of climate change on human longevity and answer the following question: What is the cost of inaction on climate change on human longevity? Our results situate climate change mortality within the broader context of human mortality and can be used to inform public health interventions and adaptation measures to prevent such futures.

Our results suggest that climate change could emerge as a potent public health threat in the 21st century. Mitigation strategies (ie reductions in greenhouse gas emissions) and adaptation strategies (ie outreach efforts, retrofitting public buildings, etc.) would help prevent this new public health concern. 

# Methods and Materials

## Data

We estimate changes in life expectancy by combining three primary data sets. First, we use the projected population's by age in conjunction with projected age-specific mortality rates from the United Nations' World Population Prospects [@united2017world]. This data provides both the future populations exposed to future climate hazards and the data necessary to calculate the difference in life expectancy between baseline, projected mortality and mortality due to climate change. Second, we use projections of excess mortality due to meteorological hazards related to climate change from Forzieri et al. [-@forzieri2017increasing]. This data provides the number of projected deaths to climate change by the end of the century. And third, we use cause-of-death distributions from the Global Burden of Disease study (GBD) [@GBD;@wang2012age] to distribute Forzieri et al's [-@forzieri2017increasing] projected death totals. 

Forzieri et al. [-@forzieri2017increasing] combine data from the International Disaster Database (EMCAT) and the Natural Catastrophe Statistics tool (NatCatSERVICE) to create exposure and fatality statistics related to six weather-related hazards -- heatwaves, cold waves, droughts, wildfires, river and coastal floods, and windstorms -- and estimate excess mortality associated with those hazards during a baseline period of 1981-2010 in 28 European countries. They only considered fatalities *directly attributable* to the event and *exclude* increased deaths from common causes that were observed to rise during the event, such as cardiovascular or respiratory deaths[^1]. While the disaster data are not standardized across country, Forzieri et al. imputed data in incomplete time periods and countries. This imputation could mask spatial variability at the sub-national level, but here we use their country-aggregated results. They downscaled these exposures rates to 1km grid cells and integrated them into small-area demographic projections for Europe out to 2100 that correspond to a middle-of-the-road socioeconomic scenario (SSP2 with medium fertility, medium mortality, medium migration, and the Global Education Trend education scenario [@jiang2014internal;@samir2017human;@o2017roads]) and based their projected mortality on extrapolations of their calculated exposures. Their results represent business-as-usual climate change and human development without incorporating potential adaptations to climate change. They found a potential 150,000+ climate change related fatalities per year by the mid 2080s with climate change contributing to 90% of mortality rise as opposed to population changes. The number of future anticipated deaths are available in their Supplementary Table S8 [@forzieri2017increasing] and provide the magnitude of deaths for our study.

[^1]: "The fatality data from the two databases considered is likely to not include increased deaths from common causes that were observed to rise at the population level but for which individual deaths could not be attributed to the event. For example,... increased risk of cardiovascular and respiratory deaths ... may be severely under-reported in the EMDAT and NatCatSERVICE." [see @forzieri2017increasing, pp. 9, *Supplementary Materials*].


<!-- # ```{r figure-GBD-death-distribution, echo= FALSE, message = FALSE, warning = FALSE, fig.cap= paste("**Mortality distribution of Heat-related mortality in 28 European countries**. The thick black line is the average\\label{suppfig}")} -->
<!-- #  -->
<!-- # ``` -->

The Forzieri et al. [-@forzieri2017increasing] data contain only mortality totals and report no age detail, precluding a direct conversion of excess mortality to life expectancy. To convert excess mortality to life expectancy, we allocate excess multi-hazard mortality for each country based on the observed age-specific mortality schedule for environmental heat and cold exposure deaths from the GBD [@GBD;@wang2012age]. Forzieri et al. found that extreme heat (as opposed to other five weather-related hazards) account for 99% of the anticipated excess mortality by the end of the century. The GBD is the most comprehensive worldwide data set of epidemiological data and provides cause-specific mortality by age/sex/geography/year for 249 causes of death in 195 countries and territories. We gather data on age-specific mortality rates ($m_x$) from environmental heat and cold exposure (cause C.2.9) for the period 2006-2015 for each country in the study for age groups $x=0,1,5,10,...95$. These data provide a mortality schedule to fit the projected climate change excess mortality from Forzieri et al [@forzieri2017increasing]. We assume that the age-specific mortality schedules observed between 2006 and 2015 due to environmental heat and cold exposure in the GBD are likely to remain unchanged in the future. We only use the GBD to distribute the Forzieri et al. deaths by age[^2].

[^2]: GBD data can be retrieved here: <http://ghdx.healthdata.org/gbd-results-tool?params=gbd-api-2017-permalink/b235b96d60b3b5f06f3e080974c415f3>.

With the Forzieri et al data and the GBD we estimate the anticipated age-specific mortality schedules due to climate change in thirty-one European countries but we cannot compare life expectancy differentials in the absence of Climate Change using those two data sets alone. To create baseline mortality schedules and life expectancy, we gather data from the United Nations' World Population Prospects (WPP-2017) [@united2017world] for the corresponding European countries in the period 2080-2085 (the mid-point of the period Forzieri et al. report their results [-@forzieri2017increasing]). We use the UN abridged life table data for ages $x=0,1,5,10,...85$ containing the necessary mortality rates($_nm_x$), the average number of years lived in the age-interval ($_na_x$), and populations ($_nP_x$) for populations aged $x$ to $x+n$.

<!-- We use the following data elements from the WPP-2017 data: -->

<!-- \begin{enumerate} -->
<!-- \item $_nP_x$ or the population in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$. -->
<!-- \item $_na_x$ or the average length of survival between ages $x$ and $x+n$. -->
<!-- \item $_nd_x$ or the number of deaths in age group $x$ for $n$-year intervals from exact age $x$ to $x+n$. -->
<!-- \end{enumerate} -->

## Methods

  We derive a number of variables to accomplish our analysis. First, we use the C.2.9 mortality rates ($_nm_x$) from the GBD for ages $x=0,1,5,10,...,95$ with the projected populations aged $x$ to $x+n$ from the UN data in the period 2080-2085 ($_nP_x$) in country $i$ to calculate an un-scaled projection of deaths due to C.2.9 ($_nD_{x,i}=_nm_{xi}^{GBD} \cdot _nP_x$). We then abridge $_nD_{xi}$ to ages 85+ to match the open-ended age interval found in the UN life tables. With abridged baseline heat/cold deaths, we proportionally rescale $_nD_{x,i}$ to sum to the projected excess mortality for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) for each individual country from Forzieri et al [@forzieri2017increasing] ($_n\hat{D}_{x,i}$) and calculate new $_nm_{xs}^{FORZ}$. $BASE$ refers to the baseline mortality rates projected by the UN for the period 2080-2085 which do not account for climate change. These calculations combine to create a projected age-specific mortality rates based on Forzieri et. al's excess mortality.
  
To calculate life expectancy differentials, we add the age-specific Forzieri rates $_nm_{xs}^{FORZ}$ to the projected age-specific mortality rates found in the published UN life tables $_nm_x$ to create a projected $\hat{m}_x=m_x + m_x^{FORZ}$. We then calculate life table components $_nq_x$, $_nd_x$,  $l_x$, $L_x$, $T_x$, and ultimately $e_x$ identically for each scenario $s$ using standard procedures, substituting $_nm_x$ with $_n\hat{m}_x$ where needed:

<!-- \begin{equation} -->
<!-- _nq_{x,i,s} = \frac{_nm_{x,i,s} \cdot  n}{1+(n-_na_{x,i}) \cdot _nm_{x,i,s}} -->
<!-- \end{equation} -->


<!-- We calculated each additional life table value identically for each scenario $s$ using standard life table equations: -->

$$_nq_{x,i,s} = \frac{_nm_{x,i,s}\cdot  n}{1+(n-_na_{x,i}) \cdot _nm_{x,i,s}}$$
$$_nd_{x,i,s} = _nl_{x,i,s} \cdot _nq_{x,i,s}$$
$$l_{x,i,s} = _nl_{x-1,i,s} - _nd_{x-1,i,s}$$
$$L_{x,i,s} = n \cdot _nl_{x,i,s} + ((n-_na_{x,i,s}) \cdot _nl_{x,i,s})$$
$$T_{x,i,s} = _nL_{x,i,s} + _nT_{x+1,i,s}$$
$$e_{x,i,s} = _nT_{x,i,s} / _nl_{x,i,s}$$

We calculate projected life expectancy reductions due to climate change related mortality by subtracting $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated]. 

<!-- the differences between baseline life expectancy and life expectancy inclusive of climate change related mortality by subtracting $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].  -->

<!-- To determine the differences between life expectancy compared to the baseline, we simply subtract $e_{x,i,s}$ (for scenario $LOW$, $MID$, and $HIGH$) from $e_{x,i,BASE}$ as described by [@beltran2008integrated].  -->

  
<!--   We then proportionally scaled $D_{x,i}$ to equal to projected excess mortality in Forzieri et. al for each individual country and calculated new $_nm_x^{FORZ}$ values. -->

<!--   We then anticipated number of deaths due to environmental heat and cold exposure in the period 2080-2085 and then proportionally scale these deaths to equal the Forzieri et al projected Climate Change deaths for the same period for each country $i$ such that -->

<!-- $$_n\hat{D}_{xi}=_nm_{xi}^{GBD} \cdot _nP_x / \sum_{\alpha=0}^{95}{_nm_{xi}^{GBD} \cdot _nP_x}$$ {#eq:eqn1} -->

<!-- The left side of equation (1) calculates the un-scaled climate change deaths in country $i$ for populations aged $x$ to $x+n$ by multiplying the underlying mortality rate for those age groups from the GBD due to COD C.2.9 ($_nm_{xi}^{GBD}$) by the populations aged $x$ to $x+n$ from the UN ($_nP_x$). The right side of equation (1) simply -->
<!-- . We then aggregate $_n\hat{D}_{xi}$ for ages 85+ to match the open-ended age interval found in the UN life tables. Finally, we calculate a new $_nm_x$ value ($_n\hat{m}_x$) representing the anticipated mortality-rates due to climate change.  -->

<!-- To compare $e_0$ in the period 2080-2085 with and without expected Climate Change mortality,  -->

<!-- abridged the life table data from the HMD from ages 100+ to 80+ to conform to the GBD cause-specific mortality schedules. -->

<!-- From the GBD data, we derived variable $t_{x,i}$ as the proportion of deaths, $D$, from each age group $x$ in each country $i$ ($_nt_{x,i}=D_{x,i}^{GBD}/\sum_{\alpha=0}^{80}{_nD_{\alpha,i}^{GBD}}$). -->

<!-- We then derive $m_x$ rates for each age group $x$ in each country $i$ for each scenario $s$ ($BASE$, $LOW$, $MID$, $HIGH$) from Forzieri et al [@forzieri2017increasing] and the WPP-2017. Scenarios $LOW$, $MID$, and $HIGH$ are calculated as the underlying mortality rate projected by the UN ($_nm_x$) plus the mortality rate due to climate change ($_n\hat{m}_x$).  -->


## Reproducible Research
All data and code necessary to reproduce the reported results are licensed under the CC-BY-4.0 license and are publicly available in a replication repository located at https://osf.io/e43sv/?view_only=f01c29b95a27405b806345e04ced62a8. 


# Results
We find that climate change could alter life expectancy by `r round(mean(lt_lancet$DIF_MID),2)` years (-2.8 months (`r round(mean(lt_lancet$DIF_LOW),2)` to `r round(mean(lt_lancet$DIF_HIGH),2)` years)) in the average European country (**\autoref{figure1}**) by 2100. This reduction is comparable to mortality due to influenza and pneumonia [@arias2013united] in the United States for the year 2000. Although the average European country could see a change of `r round(mean(lt_lancet$DIF_MID),2)` years, several countries are likely to experience considerably greater reductions in life expectancy. Spain, for example, could see a reduction of up to `r round(lt_lancet$DIF_MID[which(lt_lancet$CNTRY == "ESP")],2)*-1` years (`r round(lt_lancet$DIF_LOW[which(lt_lancet$CNTRY == "ESP")],2)*-1` - `r round(lt_lancet$DIF_HIGH[which(lt_lancet$CNTRY == "ESP")],2)*-1`).

```{r figure-lollipop, echo = FALSE, fig.cap=paste("Change in life expectancy at birth ($e_0$) due to business-as-usual climate change in the 2080s compared to the projected $e_0$. We report changes in life expectancies due to climate change for thirty-one European countries. The central values represent the ensemble median while the stems represent the upper and lower bounds of the inter-model climate variability.\\label{figure1}")}

```

Our results also suggest climate change mortality differentials are likely to unfold along highly uneven geographies (\autoref{map}). Whereas many Northern European countries could experience negligible impacts on life expectancy, seven European countries could see life expectancy changes in excess of -0.5 years (Spain, Italy, Portugal, France, Slovenia, Croatia, and Luxembourg). This group of countries could see life expectancy changes of more than -1.0 years if climate change hazards are more intense than anticipated. In some countries, climate change could thus become a bigger killer than trachea, bronchus, and lung malignant neoplasms (-0.85 years), acute myocardial infarction (-0.87 years), or all accident related mortality (-0.84 years) [@arias2013united] by the end of the century.


```{r  figure-map, echo = FALSE, warning= FALSE, message = FALSE, fig.cap=paste("Estimated reduction in life expectancy at birth ($e_0$) by the 2080s under the $MID$ scenario.\\label{map}")}
```

<!-- Unfortunately, the International Classification of Diseases (ICD-10) does not list climate change as a cause of death. The Forzieri et al estimate fatalaities directly attributable to climate events allowing us to compare climate change related mortality to other CODs. By comparing the ancitipated death rates attributate to climate change related events with other common CODs, the significance of the public health risk associated with climate change comes more clear. **\autoref{Table1}** reports the rankings of Age Standardized Death Rates (ASDR) for the leading causes of death in the ten European countries with the highest climate change related ASDRs (ASDR's for other CODs come from Eurostat (online data code: hlth_cd_asdr2)).  If climate change related mortality were its own COD it would be a top five killer in many European countries. For many countries, only combined CODs, such as all circulatory diseases or all cancers, exhibit higher ASDRs than climatic hazards. Additionally in Spain, Luxembourg, and France, climate change could become the number one killer if climate hazards unfold more severely. -->

<!-- \begin{sidewaystable} -->

<!-- \caption{\textbf{Rankings of major causes of death in select European Countries based on a standardized death rate (per 100,000 inhabitants).} We compare climate change related mortality to major causes of death (COD) in thirty-one European Countries using age standardized death rates. Climate change is not a COD but comparisons here group climate change related mortality directly attributable to events making these comparisons possible. If climate change related mortality were it own COD, it is projected to be the second leading COD in Spain and Luxembourg. Data for other causes of death comes from Eurostat.} \label{Table1} -->
<!-- \rowcolors{2}{gray!25}{white} -->
<!-- \footnotesize -->
<!-- \resizebox{\linewidth}{!}{% -->
<!-- \begin{tabularx}{\textwidth}{l|XXXXXXXXX} -->
<!--   \hline -->

<!--    \rowcolor{white} -->
<!--  Country & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 \\  -->
<!-- \hline -->
<!-- Spain & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 91.7\} & makecell\{Heart\textbackslash{}Disease\textbackslash{} 68.2\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 48.5\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 47.8\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 44.99\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 33.6\} & makecell\{Suicide\textbackslash{} 8.2\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 4.3\}\\ -->
<!-- \hline -->
<!-- Luxembourg & makecell\{Heart\textbackslash{}Disease\textbackslash{} 80.3\} & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 63.8\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 59.6\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 38\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 25.5\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 18.52\} & makecell\{Suicide\textbackslash{} 13.4\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 6\}\\ -->
<!-- \hline -->
<!-- Italy & makecell\{Heart\textbackslash{}Disease\textbackslash{} 98.3\} & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 58.3\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 49.4\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 34.3\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 27\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 23.48\} & makecell\{Suicide\textbackslash{} 6.3\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 5.6\}\\ -->
<!-- \hline -->
<!-- France & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 52\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 50.2\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 50.1\} & makecell\{Heart\textbackslash{}Disease\textbackslash{} 49.3\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 26.1\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 18.7\} & makecell\{Suicide\textbackslash{} 14.1\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 5.1\}\\ -->
<!-- \hline -->
<!-- Portugal & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 116.7\} & makecell\{Heart\textbackslash{}Disease\textbackslash{} 69.6\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 36.4\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 35\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 32.8\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 22.2\} & makecell\{Suicide\textbackslash{} 11.3\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 7.8\}\\ -->
<!-- \hline -->
<!-- Ireland & makecell\{Heart\textbackslash{}Disease\textbackslash{} 147.5\} & makecell\{Respiratory\textbackslash{}Diseases\textbackslash{} 125.9\} & makecell\{Lung\textbackslash{}Cancer\textbackslash{} 61.5\} & makecell\{Dis. of the\textbackslash{}Nervous Sys\textbackslash{} 48.7\} & makecell\{Colorectal\textbackslash{}Cancer\textbackslash{} 32.4\} & cellcolor\{blue!25\}makecell\{cellcolor\{blue!25\}Climate\textbackslash{}cellcolor\{blue!25\}Change\textbackslash{}cellcolor\{blue!25\} 11.61\} & makecell\{Suicide\textbackslash{} 11\} & makecell\{Transport\textbackslash{}Accidents\textbackslash{} 4\}\\ -->
<!-- \hline -->
<!-- \end{tabularx} -->
<!-- } -->

<!-- \end{sidewaystable} -->

# Discussion
In this article, we demonstrate the impact climate change could have on life expectancy at birth due to extreme weather events in thirty-one European countries. Previous studies on climate change and excess mortality potentially miscommunicate the impact climate change could have on human mortality. Contrary to excess mortality estimates, life expectancy is routinely used as a primary metric for communicating overall health outcomes and enjoys widespread use by major international organizations [@world2015world;@marmot2012building;@salomon2012healthy]. Life expectancy and its derivatives are the recommended metrics for population health [@parrish2010peer]. Additionally, it connects mortality estimates into intuitively understandable metrics, translating global estimates of mortality into personally relatable outcomes. Our work reveals the extent to which climate change could reduce the average person's longevity by the end of the century, expanding our understanding of climate change and public health; thus linking two of the major areas in current developmental and sustainability discussion at both national and international levels [@abel2016meeting]. 


Without adaptation measures, our results suggest climate change could emerge as a significant new mortality vector and could pose a major public health threat for some European countries by the end of the century, echoing previous findings [@forzieri2017increasing;@patz2005impact]. Life expectancy has steadily risen across the world for the last century [@gerland2014world] and our results suggest that climate change alone could slow these trends in some countries. We expect one European country to see life expectancy reductions of more than one year under the middle scenario (Spain), but if climate change has a greater impact on mortality than anticipated, four European countries (Spain, Italy, France, Luxembourg) could see life expectancy reductions of more than one year with Spain experiencing a reduction of nearly two. These findings highlight the hyperlocalized impacts of climate change [@kendon2014heavier;@rosenzweig2010cities;@forzieri2017increasing]. 

Reductions such as these should not be taken lightly. Many of the children born today are likely to still be alive by the end of the century and will be in the age groups (aged 65+) most threatened by the biggest mortality risk associated with climate change [@keatinge2000heat] -- extreme heat. 
<!-- If climate change unfolds as a more aggressive mortality vector, only all circulatory diseases combined or all cancers combined would contribute more to mortality rates in numerous European countries.  -->
Climate change could thus be one of the most aggressive new mortality vectors to emerge over the last quarter-century, representing a major threat to public health in many parts of Europe.

Prospective studies on the emerging threat from climate change rely on linking contemporary mortality with future mortality. However, climate change could reshape future mortality through other causes of death.  Climate change affects health behaviors that in turn increase mortality risk through increased alcohol and substance abuse, violent behavior, insecurity, increase in post-traumatic stress due to weather-related trauma, increase in stress due to climate change and schizophrenia, increase in the use of medications that reduce the ability to perspire and sweat, etc. [@patz2005impact] The International Classifications of Diseases and Related Health Problems (ICD) does not contain ``climate change" as an official cause of death, so we can only speculate that the impact of climate change could be larger than reported here. Although we do not model these potential impacts, our results could thus be considered conservative.
    
These results should also be considered conservative when compared to the broader impact of climate change on human longevity. The disaster databases that Forzieri et al [@forzieri2017increasing] use in generating their excess mortality estimates probably, though the disaster databases are unclear, only account for the deaths certified as *directly* caused by these hazards and are unlikely to capture the overall numbers associated with these deaths. If the certification of deaths due to weather extremes is similar to the certification in the present, than our results and those of Forzieri et al reflect the excess mortality directly attributable to climatic extremes. Despite this limitation, the impact of these extremes on $e_0$ is considerable, even if conservative.    
    
We also share the concerns of Lee et al. [@lee2017comprehensive] concerning the business-as-usual climatic assumptions. It is likely that many countries and communities will deploy a wide variety of adaptation measures [@haines2006climate;@kovats2003methods;@ebi2006approach]. These adaptation measures rely on accurate information about the potential mortality vectors. Our models and those produced by Forzieri et al. [@forzieri2017increasing] present plausible scenarios on the potential impact of climate change on human mortality and provide crucial information to public health officials, national governments, and international organizations. The time frames associated with climate change allow ample time for this potential health crisis to be averted.

Finally, we would like to point out that future research should not only transform excess mortality into life expectancy decrements. Given the influence of climate change in diseases and causes of death, it is imperative to quantify the extent to which climate change will derive in increasing costs for health care systems in these countries. The health care structures are being taxed by population aging [@rechel2009can] and rising health care costs, yet it remains unclear how climate change will exacerbate these pressures.


